Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-16 paper/ms.tex
words about parameterization
adding notation for the Lorentzian limit and a notebook checking the parameters
Added computation of SHO, ACF of spot models

% The approach we explore in this paper is based upon a method developed by Press \&
% Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
% They also indicated that a combination of two exponential kernels would have a simliar scaling,
% but they did not demonstrate how to extend this to more than two kernel components.

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
% Rybicki-Press' (\texttt{GenRP}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

@@

% \section{Outline}
% \begin{enumerate}
%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
%\item Assumptions:\\
%  a). Stationary noise (although this can be broken).\\
%  b). Gaussian
%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
%   curves; Solar curves; short-cadence.
%\item Solution(s):\\
%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
%  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
%  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
%\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
%   $O(N*p^2)$, where p is number of Lorentzians.
%  - Sum of Lorentzians is a good description of stellar variability:
%     - granulation has expoential (or sums of exponential) noise
%     - asteroseismic obeys sums of Lorentzians
%     - periodic variability due to spots/plages/etc. can also be
%       approximated by Lorentzian(s)
%\item Generalized Ambikasarn for complex coefficients (Hermitian).
% \item Performance:\\
%   a). Show $O(N*p^2)$ scaling\\
%   b). Show comparison to HODLR.\\
%   c). Show comparison to wavelet (ala Carter \& Winn).\\
%   d). Compare with CARMA Kallman filter(?).
% \item Various methods/formulae:\\
%   a). Likelihood computation.\\
%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
%   c). Inferring particular components.\\
%   d). Computing derivatives of GPs (as used in RV methods).\\
%   e). Forecasting/interpolating.\\
%   f). Inferring confidence intervals.\\
%   g). Derivative of likelihood function with respect to kernel parameters.\\
%   h). Non-stationary GP (can the coefficients vary with time?).\\
%   i). Multi-band (each element of extended matrix becomes a matrix).
%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
% \item Generalization to non-stationary noise.
% \item Example application: TYC 3559\\
%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
%   b). Analysis of solar data.
%   c). CO2 Mauna Kea
%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
%   e). TBD Forecasting/interpolating.\\
%   f). TBD Inferring confidence intervals.\\
%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
%   i). TBD Multi-band (each element of extended matrix becomes a matrix).

@@

Now, the damped harmonic kernel components may be rewritten as:
\begin{equation}
 a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
\end{equation}
where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran2015}.
The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
but otherwise the mathematical form of the extended matrix is identical, but with twice as many
exponential kernel components as the real case.

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.    Within astrophysics, correlated noise (due to the environment,
detector, or modeling uncertainty) may be present in
gravitational wave time series, and so Gaussian processes may be a way to address
this problem \citep{Moore2016}.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,1995PhRvL..74.1060R}.
This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.

@@

Here we summarize \citet{Ambikasaran2015}, and then point to appendix
for how to do this with oscillating kernels.

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

@@

For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $p$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
to be the real values of $\beta$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
\end{equation}
for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
in \citet{Ambikasaran2015}.

Equation (60) in \citet{Ambikasaran2015} has real and imaginary components:
\begin{eqnarray} \label{Amb60}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.

Finally, equation (59) in \citet{Ambikasaran2015} has real and imaginary components:
\begin{align} \label{Amb59}
-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) in \citet{Ambikasaran2015} as well for these, which reduces the total number
of equations to solve.

@@

The matrix has a similar structure as the real case constructed by \citep{Ambikasaran2015}, but the
imaginary and real components mix (this is due to the representation of imaginary numbers as
$2\times 2$ matrices).  Figure \ref{matrix_structure} shows an example of the structure of the extended
matrix for two Lorentzians with non-zero $\omega$ values.  This is analagous to the real case with
$p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran2015}.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{./twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=1.0]{./colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
the example shown in \citet{Ambikasaran2015}.}
\label{matrix_structure}
\end{figure}

>>>>

% The approach we explore in this paper is based upon a method developed by Press \&
% Rybicki twenty years ago \citep{Rybicki:1995}.  They observed that a correlation
% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
% They also indicated that a combination of two exponential kernels would have a simliar scaling,
% but they did not demonstrate how to extend this to more than two kernel components.

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
% Rybicki-Press' (\texttt{GenRP}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

@@

% \section{Outline}
% \begin{enumerate}
%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
%\item Assumptions:\\
%  a). Stationary noise (although this can be broken).\\
%  b). Gaussian
%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
%   curves; Solar curves; short-cadence.
%\item Solution(s):\\
%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
%  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
%  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
%\item Solution: Ambikasaran \citep{Ambikasaran:2015} shows how sum of exponential kernels can be solved in order
%   $O(N*p^2)$, where p is number of Lorentzians.
%  - Sum of Lorentzians is a good description of stellar variability:
%     - granulation has expoential (or sums of exponential) noise
%     - asteroseismic obeys sums of Lorentzians
%     - periodic variability due to spots/plages/etc. can also be
%       approximated by Lorentzian(s)
%\item Generalized Ambikasarn for complex coefficients (Hermitian).
% \item Performance:\\
%   a). Show $O(N*p^2)$ scaling\\
%   b). Show comparison to HODLR.\\
%   c). Show comparison to wavelet (ala Carter \& Winn).\\
%   d). Compare with CARMA Kallman filter(?).
% \item Various methods/formulae:\\
%   a). Likelihood computation.\\
%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
%   c). Inferring particular components.\\
%   d). Computing derivatives of GPs (as used in RV methods).\\
%   e). Forecasting/interpolating.\\
%   f). Inferring confidence intervals.\\
%   g). Derivative of likelihood function with respect to kernel parameters.\\
%   h). Non-stationary GP (can the coefficients vary with time?).\\
%   i). Multi-band (each element of extended matrix becomes a matrix).
%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
% \item Generalization to non-stationary noise.
% \item Example application: TYC 3559\\
%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
%   b). Analysis of solar data.
%   c). CO2 Mauna Kea
%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
%   e). TBD Forecasting/interpolating.\\
%   f). TBD Inferring confidence intervals.\\
%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
%   i). TBD Multi-band (each element of extended matrix becomes a matrix).

@@

Now, the damped harmonic kernel components may be rewritten as:
\begin{equation}
 a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
\end{equation}
where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
but otherwise the mathematical form of the extended matrix is identical, but with twice as many
exponential kernel components as the real case.

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.    Within astrophysics, correlated noise (due to the environment,
detector, or modeling uncertainty) may be present in
gravitational wave time series, and so Gaussian processes may be a way to address
this problem \citep{Moore2016}.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.

@@

Here we summarize \citet{Ambikasaran:2015}, and then point to appendix
for how to do this with oscillating kernels.

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

@@

For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $p$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
to be the real values of $\beta$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
\end{equation}
for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
in \citet{Ambikasaran:2015}.

Equation (60) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{eqnarray} \label{Amb60}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.

Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
\begin{align} \label{Amb59}
-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
of equations to solve.

@@

The matrix has a similar structure as the real case constructed by \citep{Ambikasaran:2015}, but the
imaginary and real components mix (this is due to the representation of imaginary numbers as
$2\times 2$ matrices).  Figure \ref{matrix_structure} shows an example of the structure of the extended
matrix for two Lorentzians with non-zero $\omega$ values.  This is analagous to the real case with
$p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran:2015}.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{./twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=1.0]{./colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
the example shown in \citet{Ambikasaran:2015}.}
\label{matrix_structure}
\end{figure}
